---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Manipulating data

At this point, you can set up an R project, import a data file into RStudio, and quickly explore the dataset's structure, which is typically the first step in any analysis. If the dataset is well-structured, the next crucial step is to manipulate the data in R to obtain the information we need. Often, we are not interested in the entire dataset; instead, we focus on selecting specific variables, analyzing certain groups, and observing how these variables vary across those groups

With R, there are several ways to perform these operations. Firstly, you can use R's base functions, without having to load a specific package. An alternative approach is based on the `tidyverse`, the package suite developed by [Hadley Wickham](https://en.wikipedia.org/wiki/Hadley_Wickham) with a special syntax. Here, we'll concentrate on the `tidyverse` functions, which I feel are easier and more intuitive to use.

## Introduction to dplyr functions

To manipulate data, we will use a package from the tidyverse called `dplyr`, comprising a comprehensive suite of exceptionally useful functions. To familiarize ourselves with its usage, we will explore the [Quality of Government Environmental Indicators](gu.se/en/quality-government/qog-data/data-downloads/environmental-indicators-dataset) dataset. To do this, we will need first to load the tidyverse and import the data with the `haven` package.

```{r}
#| warning: false
# Load the tidyverse

library(tidyverse) 
library(haven)

qog <- read_dta("data/qog_env.dta") 

head(qog)

```

### Count the observations of groups with `count()`

One basic operation in data manipulation is to count the number of observations of different values in a variable. The `dplyr::count()` function is used for this purpose.

In Exercise 1, we observed that the qog dataset has a country-year structure. For each country, there is a series of indicators, with one value per year. The dataset includes a `year` variable and a `cname` variable representing the country.

We might want to know how many country observations we have for each year. To do this, we can use `count()` with the `year` variable, that will give us the number of observations for each year. If we look at the results, we see that the dataset has observations from 1946 to 2020 with an increasing number of countries in the dataset over time.

```{r}
count(qog, year)
table(qog$year) # Alternative way to do it
```

If we want to cross the `year` and `cname` variables, we can use the `count()` function with both variables. This will give us the number of observations for each year and country. In this case, we see that we have only one observation for each country-year combination.

```{r}
count(qog, year, cname)
```

### `filter()` observations

We may want to use only a subset of countries and get a smaller versions of the dataset. And the tidyverse has one for this, which is called `filter()` We want to select specific rows/observations of the dataset based on a specific conditions. To filter, we will use extensively boolean operators.

| Operator | Description           |
|----------|-----------------------|
| `==`     | equal                 |
| `!=`     | not equal             |
| `<`      | less than             |
| `<=`     | less than or equal    |
| `>`      | greater than          |
| `>=`     | greater than or equal |
| `&`      | and                   |
| `|`      | or                    |
| `!`      | not                   |
| `%in%`   | in the set            |

For instance, I might be only interested in environmental indicators for France. In that case I could filter the whole dataset with only the observations that has France as country variable.

```{r}
filter(qog, cname == "France")

# If I want to keep all the countries except France
filter(qog, cname != "France") 

# Filter for multiple countries
filter(qog, cname %in% c("France", "Norway", "Spain", "Denmark"))

# Filter for countries and year and save in a new object ! 
qog_subset <- filter(qog, cname %in% c("France", "Norway", "Spain", "Denmark"), year > 1990)

count(qog_subset, cname)
```

Often however, we need several operations to be done together. To do so with the tidyverse functions, we can use something that is called the pipe : `|>`. You will quickly understand what the pipe is, we use it to chain instructions just as in a recipe. Depending on the version of R, you might also use/see this `%>%`. The shortcut for it is `Ctrl + Shift + M` or `Cmd + Shift + M` on Mac.

Let's say we want to count the number of observations we have on the variable `wvs_pedp` that measure the share of the population participating in environmental protests for Sweden. We can use the pipe to do it by chaining the `filter()` and `count()` functions rather than doing it separately in different steps.

```{r}
qog |> 
  # Keep only countries 
  filter(cname == "Sweden") |> 
  # Count the number of observations for each country
  count(wvs_pedp) # Use Sort = TRUE if you want to sort the results
```

### Your turn

Count the countries and year where the share of the population participating in environmental protests was greater than 10 (measured by the `wvs_pedp` variable).

## `select()` variables

While `filter()` is useful for keeping only certain groups of rows, `select()` is used, as its name suggests, to select certain variables (columns) from our dataframe.

Let's say I'm interested in CO2 emissions per capita, their overall level and how they vary over time and by country. For this, I would need only three variables that are present in the dataset: `year`, `cname` and the variable `wdi_co2` which comes from the [World Bank's World Development Indicators](https://databank.worldbank.org/source/world-development-indicators). To do this, I use `select()` and just specify which variables of the dataset I want to select.

```{r}
qog_subset |> 
  select(cname, year,  wdi_co2)
```

## Calculating statistics by group with `group_by()` and `summarise()`

To compute a first set of descriptive statistics, we could look at the central indicators of the co2 emissions variable (`wdi_co2`).

```{r}

mean(qog_subset$wdi_co2, na.rm = TRUE) # Compute the mean
median(qog_subset$wdi_co2, na.rm = TRUE) # Compute the median
sd(qog_subset$wdi_co2, na.rm = TRUE) # Compute the standard deviation
```

However, we are often interested in how these indicators vary across groups such as country or year. To do so, use the `summarise()` function in R that allows you to compute new variables by groups. To choose grouping variable, we use first `group_by()` where we specify for which group we want to compute something. Here I use this function to compute the mean, the median, the standard deviation and the first and third quartile of co2 emissions per capita for each country. This gives us a new tibble with all of the summary information.

```{r}
qog_stats <- qog |> 
   group_by(cname) |> 
  summarise(mean_co2 = mean(wdi_co2, na.rm = TRUE), 
            median_co2 = median(wdi_co2, na.rm = TRUE),
            sd_co2 = sd(wdi_co2, na.rm = TRUE))

qog_stats
```

It is also possible to sort the results to gain a clearer idea of which countries have the highest average CO2 emissions per capita. To achieve this, we utilize the `arrange()` function, specifying the variable by which we intend to sort.

```{r}
qog_stats |> 
  arrange(-median_co2)

qog_stats |> 
  arrange(mean_co2)
```

## Your turn

When were the CO2 emissions per capita the lowest in Saudi Arabia?

### Mutate

We can also create new variables based on other ones with `mutate()`. Let's say I want to compute the growth rate of co2 emissions per capita every year. For this, I will use `mutate()` to create new variables.

```{r}
qog |> 
  select(cname, year, wdi_fossil) |> 
  mutate(dataset = "QOG dataset")
```

To compute the growth rate, we need for each year the emissions per capita of that year and those from the previous year, which I access with `lag()`.

```{r}

qog_co2_growth <- qog |>
  filter(cname == "France") |>
  select(cname, year, wdi_co2) |> 
  mutate(co2_lag = lag(wdi_co2),
        co2_growth = (wdi_co2 - co2_lag) / co2_lag*100)

qog_co2_growth 
```

## Recode values with `case_when`

`Mutate()` is also widely used to recode variable. Here we create a new `trajectory` variable and we use `case_when()` to recode whether there co2_emissions are growing or not for a given year.

```{r}
qog_co2_growth <- qog_co2_growth |> 
  filter(cname == "France") |>
  mutate(
    trajectory = case_when(
      co2_growth > 0 ~ "Bad",
      co2_growth < 0 ~ "Good"
    )
  )
qog_co2_growth

qog_co2_growth |> 
  count(trajectory)
```

### Your turn !

Create a century variable that indicates whether the year is in the 20th or 21st century. Find the countries that have the highest total number of deaths due to natural disasters for each century, as measured by the `emdat_ndeath` variable.
