## Going further

To get nice tables when you render your quarto document, you can use the `kableExtra::kable()` function.

```{r}
#| output: asis 

library(kableExtra)


qog_stats |> kable()
```

The dyplr package also allows to perform the same operation on multiple variables at once, which can be useful when you have a lot of variables to work with. For instance here, I want to remove the `wdi_` prefix from all the variables that start with `wdi` and use the `rename_with()` function to do so.

```{r}
# More complex ! 

qog_renamed <-  qog_wdi |> 
  # Remove wdi_ from all of the variables names that starts with wdi
  rename_with(~ str_remove(.x, "wdi_"), starts_with("wdi")) 

qog_renamed |> colnames()

```

Similarly, we can use the `across()` function to apply the same function to multiple variables at once. Here I want to compute the mean of all the variables that start with `wdi`.

```{r}
qog_growth <- qog |> 
  group_by(cname) |> 
  summarise(across(c(wdi_co2, wdi_fossil),~ mean(.x, na.rm = T)))

qog_growth
```

```{r}
qog_subset |> slice_max(wdi_co2, n = 5) # Get the n max results of a variable
qog_subset |> slice_min(wdi_co2, n = 3) # Get the n min results of a variable
```

The `select()` functions offer a lot of possibilities to select variable based on certain patterns by adding functions inside such as `starts_with()`, `ends_with()`, `contains()`. You can learn more by typing `?select`.

```{r}
colnames(qog_subset)

qog_wdi <- qog_subset |> 
  # Select the country, the year and every variable that starts with wdi. 
  select(cname, year, starts_with("wdi"))

# Look at which columns we have in our dataset

qog_wdi |> colnames()

# Select all of the variable that do not starts with wdi

qog_wdi |> select(-starts_with("wdi")) |> colnames()

# Select column by position (from 1 to 5)

qog_wdi |> select(1:5) |> colnames()

qog_wdi |> colnames()

qog_wdi |> 
  # Select only the variables that have numeric as type
  select_if(is.character) |> 
  # Look at which columns we have in our dataset
  colnames()
```

Here, see that the tidyverse functions are a bit intuitive that what you could do directly with R base functions. To do the same, here an example of what we would have done :

```{r}
qog[qog$cname %in% c("France"), ]
```

## Subset rows with `slice()`

You can also subset rows by their positions using `slice()`.

```{r}
qog_subset |> slice(5:7) # Subset only rows 5 to 7
qog_subset |> slice(-1) # Remove the first row
```
## Change column names with `rename()`

Occasionally, you may encounter datasets with columns named in a non-meaningful manner, such as v5, x45, or Q234, often representing variable identifiers from surveys. To enhance the understanding of the variables' meanings, it can prove beneficial to rename these columns using the rename() function. In our current dataset, the variable names are already quite informative. However, for the purpose of this illustration, let's rename the 'cname' variable as 'country', offering a more explicit label.

```{r}
qog_renamed <- qog |> 
  # Rename the variable cname as country
  rename(country = cname) 

qog_renamed |> 
  head()
```